A Graph Theory Based Approach to MDAO Problem Formulation

### 2013-03-15 graph_syntax.tex
updated introduction and started editing a couple of lower sections

\section{Graph--based Problem Formulation Syntax}
A problem formulation will be represented by nodes and edges which will be assigned certain properties and rules. There are many MDAO problem formulation concepts that we wish to represent. An analysis code should be represented by a distinct structure to distinguish between the input, output, and function evaluation, processes. The variables passed between analysis codes should be represented separately to provide insight and control over the dataflow. For example, it is important to distinguish between the case where two analysis codes are supplying the same variable to a third code, or the case where the two analysis codes are providing different variables to the third code. 
Local inputs, local outputs, and design variables(??) should be distinguished from global inputs, which are input that are used by multiple analysis codes.
Objectives and constaints should be represented as separate from local variables because they must be provided in order for the problem formulation to be valid. The coupling between sets of analysis codes is important to capture, and multi--fidelity analysis should also be allowed.

\subsection{Graph Theory Basics}
We begin by discussing the general syntax of graph theory and then expand upon this syntax to represent the data flow of an MDAO problem. The present notation is adapted from Diestel \cite{Diestel2010}.
A \emph{graph} is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, which means that the elements of $E$ are 2--element subsets of $V$. The set $V$ contains the \emph{vertices} or \emph{nodes} and the set $E$ contains the \emph{edges}.
For a \emph{directed graph*} we construct $E$ as a set of ordered pairs instead of a set of sets. Each ordered pair represents an edge starting at the node indicated by the first entry and directed to the node indicated by the second entry. Edge $e$ = $(x,y)$ may be referred to simply as $xy$ and we say $y = E(x)$. 
\begin{figure}[htb!]
	\begin{center}
	\includegraphics[width=1.5in]{images/example_directed_graph}
	\end{center}
	\vspace{-20pt}
\caption{Example directed graph.}
\label{f:example directed graph}
\end{figure}
As an example, for the directed graph shown in Fig.~\ref{f:example directed graph} we have
\begin{IEEEeqnarray*}{rCl}
V & = & \{1,2,3,4,5\}, \\
E & = & \big\{(1,2),(3,2),(1,3),(3,4)\big\}.
\end{IEEEeqnarray*}

@@

\subsection{Graph Theory Representation of a Problem Formulation}
In order to represent the listed problem formulation concepts with a graph, we assign properties and rules to nodes and edges.
The first property assigned to nodes and edges is the \emph{type}. \\
The possible node types are:
\begin{description}
\item[variable node] The variable node represents passing of data.
\item[model node] The model node represents the handling of data.
\end{description}
The possible edge types are:
\begin{description}
\item[fixed edge] A fixed edge may not be removed.
\item[free edge] A free edge may be removed.
\end{description}
Model nodes must be used with the following rules:
\begin{itemize}
\item A model node can only have one edge directed to or from another model node.
\item A model node can only have fixed edges directed in or out.
\item A model node must have at least one edge directed in and at least one edge directed out.
\end{itemize}
The MDO problem formulation concepts represented by nodes are given in Table \ref{t:node representation}, and the concepts represented by nodes are given in Table \ref{t:edge representation}.
% \begin{itemize}
% \item An objective or constraint is indicated by a variable node with no outgoing edges and with the incoming edges being directed from only other variable nodes.
% \item A global input is a variable node with no edges directed in and with at least two edges directed out to different variable nodes.
% \item A design variable is represented by a variable node with no edges directed in and on fixed edge directed out.
% \end{itemize}
\begin{table}[h!]
 \begin{center}
  \caption{Problem formulation concept represented by a node}
  \label{t:node representation}
  \begin{tabular}{ccc} \hline 
edges directed in & edges directed out & node representation \\ \hline
only free edges & none & objective or constraint\\
none & at least two free edges & global input \\
none & one fixed edge & design variable \\ \hline
  \end{tabular}
 \end{center}
 \vspace{-15pt}
\end{table}
\begin{table}[h!]
 \begin{center}
  \caption{Problem formulation concept represented by an edge}
  \label{t:edge representation}
  \begin{tabular}{ccc} \hline 
from node & to node & edge representation \\ \hline
variable & variable & connection/passing of a variable\\
variable & model & local input \\
model & variable & local output \\
model & model & function call \\ \hline
  \end{tabular}
 \end{center}
 \vspace{-15pt}
\end{table}

>>>>

\section{New Graph--based Syntax}
The ultimate goal of the new graph--based syntax presented here is to be able to 
fully describe the fundamental problem formulation for a complex design task, 
while still being able to accomodate the more specific case when a solution 
strategy is applied to the problem formulation. In order to achieve that goal 
the graph needs to accomodate a number of constructs from MDAO problem:

\begin{itemize}
    \item Analysis tools 
    \item Local and global variables 
    \item Coupling between analyses
    \item Design variables
    \item Objective or objectives 
    \item Local and global Constraints
    \item Iterative loops from solvers and optimizers
\end{itemize}

The following section provides an overview of graph theory basics, then describes 
in detail the different node and edge types, as well as rules governing allowable 
egdegs between nodes, used as the foundation of the new graph syntax. These basic graph 
primatives are them mapped to the MDAO constructs stated above, and used to 
build three distinct graph forms:

\begin{itemize}
    \item Maximal Connectivity
    \item Fundamental Problem Formulation 
    \item Specific Solution
\end{itemize}

Each of these forms serves a different purpose in the design process, and we 
present some transformations that can be applied to transition 
between any two of the three graph forms.

\subsection{Graph Theory Basics}
The notation used in this work is adapted from Diestel \cite{Diestel2010}. 
A \emph{graph} is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, 
which means that the elements of $E$ are 2--element subsets of $V$. The set $V$ 
contains the \emph{vertices} or \emph{nodes} and the set $E$ contains the \emph{edges}.
For a \emph{directed graph*} we construct $E$ as a set of ordered pairs instead 
of a set of sets. Each ordered pair represents an edge starting at the node 
indicated by the first entry and directed to the node indicated by the second 
entry. Edge $e$ = $(x,y)$ may be referred to simply as $xy$ and we say $y = E(x)$.

\begin{figure}[htb!]
	\begin{center}
	\includegraphics[width=1.5in]{images/example_directed_graph}
	\end{center}
	\vspace{-20pt}
\caption{Example directed graph.}
\label{f:example directed graph}
\end{figure}
As an example, for the directed graph shown in Fig.~\ref{f:example directed graph} we have
\begin{IEEEeqnarray*}{rCl}
V & = & \{1,2,3,4,5\}, \\
E & = & \big\{(1,2),(3,2),(1,3),(3,4)\big\}.
\end{IEEEeqnarray*}

@@

\subsection{MDAO Graph Primatives}
There are three node types:  
\begin{description}
\item[variable node] represents passing of data.
\item[model node] represents the handling of data.
\item[driver node] represents constrol structures capable of managing iteration
\end{description}
There are two edge types: 
\begin{description}
\item[fixed edge] A fixed edge may not be removed.
\item[free edge] A free edge may be removed.
\end{description}

A number of restrictions are places on the model node with respect to the number and 
type of edges that can be connected to it: 
\begin{enumerate}
\item A model node can only have one edge directed to or from another model node.
\item A model node can only have fixed edges directed in or out.
\item A model node must have at least one edge directed in and at least one edge directed out.
\end{enumerate}

The MDO problem formulation concepts represented by nodes are given in 
Table \ref{t:node representation}, and the concepts represented by edges are 
given in Table \ref{t:edge representation}.
% \begin{itemize}
% \item An objective or constraint is indicated by a variable node with no outgoing edges and with the incoming edges being directed from only other variable nodes.
% \item A global input is a variable node with no edges directed in and with at least two edges directed out to different variable nodes.
% \item A design variable is represented by a variable node with no edges directed in and on fixed edge directed out.
% \end{itemize}
\begin{table}[h!]
 \begin{center}
  \caption{Problem formulation concept represented by a node}
  \label{t:node representation}
  \begin{tabular}{ccc} \hline 
edges directed in & edges directed out & node representation \\ \hline
only free edges & none & objective or constraint\\
none & at least two free edges & global input \\
none & one fixed edge & design variable \\ \hline
  \end{tabular}
 \end{center}
 \vspace{-15pt}
\end{table}
\begin{table}[h!]
 \begin{center}
  \caption{Problem formulation concept represented by an edge}
  \label{t:edge representation}
  \begin{tabular}{ccc} \hline 
from node & to node & edge representation \\ \hline
variable & variable & connection/passing of a variable\\
variable & model & local input \\
model & variable & local output \\
model & model & function call \\ \hline
  \end{tabular}
 \end{center}
 \vspace{-15pt}
\end{table}

### 2013-03-15 intro.tex
updated introduction and started editing a couple of lower sections

        As the size and complexity of engineering systems grow, the time and expense for setting up 
        analysis models grow with them. Multidisciplinary Design Analysis and Optimization (MDAO)
        frameworks such as OpenMDAO\cite{Gray2012} and ModelCenter have enabled a new level of analysis tool integration 
        and paved the way for models with more analyses and increasing numbers of interdisciplinary couplings. 
        Such complex models present a distinct challenge for proper implementation of any given solution strategy.

        We propose the use of graph-based problem formulation that generally describes a
        given problem, supports algorithmic manipulation to find beneficial solutions strategies, and enables automated 
        implementation of those strategies inside an MDAO framework. In order to serve that purpose, the graph must include the following information: 
        \begin{itemize}
           \item Discipline Analyses 
           \item Discipline State Variables and Residuals
           \item Design Variables
           \item Constraints
           \item Objective or Objectives
           \item Coupling Constraints
           \item Parameters
        \end{itemize}
        Notably absent from the preceding list are any kind of solvers, optimizers, or other iterative solution finding tools. 
        At its most basic, a problem formulation includes only information about what is being sought after in a given problem or what the
        goals of a given problem are. It need not contain any information about solution paths or strategies to reach those goals.

        We define a problem specification striped down to its most essential parts to be the 
        Fundamental Problem Formulation (FPF). By definition, the FPF for any given problem will be constant regardless 
        of which MDAO framework, optimization architecture, optimization algorithm, or iterative solver is used to solve the problem.

        The graph-based syntax used in this work is based on the Design Structure Matrix (DSM). 
        This provides several key features that make it useful for working with large-scale MDAO problems. 
        It provides a rigid structure that can be easily manipulated with a wide range of well-established graph-theory algorithms 
        for the purposes of problem decomposition. It also provides a means for testing a given 
        problem formulation to check if it is the FPF, and if not, to reduce it to the FPF. Lastly, it lends itself 
        well to interacting with MDAO frameworks.

    \section{Specific vs Fundamental Problem Formulation }
        A simple, notional problem is given as follows:

@@

        Where $A$ and $B$ represent analysis tools, and $f$ and $g$ are the objective and constraint functions respectively. 
        Equation \ref{eqn:simple} makes an inherent assumption about the solution strategy for the problem. 
        Analysis $A$ outputs $z$, which is an input to $B$. Hence, $A$ should be run before $B$ with 
        the $y$ being iterated on to convergence with $y^t$. However, a different solution could be equally valid and still represent
        the same fundamental problem:

@@

        Equation \ref{eqn:simple2} differs only slightly from Eq. \ref{eqn:simple}. $A$ is now dependent on the output of $B$, 
        and $z$ will be iterated on to satisfy $g$. Now the problem can be solved by running $B$ first and then $A$.
        Since the formulations in Eqs. \ref{eqn:simple} and \ref{eqn:simple2} both describe the same problem and neither can be the
        FPF, there must be a more fundamental description of the problem that is common between them. We present the FPF as follows:

@@

    \section{Using DSM for Problem Formulation}

        As shown above, the mathematical language for specifying problem formulations is very general and can be used both for 
        fundamental and specific problem formulations. Tedford and Martins used the above mathematical syntax to specify the 
        FPF for a set of test problems and also to describe specific formulations for solving them with a 
        number of optimization architectures\cite{Tedford2009}. Their work demonstrates clearly how multiple specific 
        problem formulations can all relate back to a common FPF.

        The challenge with using this traditional mathematical syntax is that it is not easily manipulated or analyzed. 
        A number of matrix-based methods have been used successfully to translate the mathematical syntax into a more useful computational form. 
        Steward's Design Structure Matrix (DSM) is a square adjacency matrix which captures the relationship between analysis tools where off 
        diagonal elements of the matrix indicate coupling\cite{Steward1981}. Since a DSM describes a square adjacency matrix, 
        it can be represented in an equivalent directed graph where nodes represent analysis tools and 
        edges represent information dependence between those tools. The ordering of elements in a DSM can be used to indicate 
        execution order.  For more complex problems, choosing the proper order to run analysis tools is a non-trival task. 
        Rogers et. al developed DeMAID to manipulate a DSM to find an ordering for analysis tools that 
        reduces the cost of solving highly coupled systems\cite{Rogers1996}. This re-ordering yields multiple specific problem 
        formulations which all solve the same FPF. In other words, manipulation of the DSM does not fundamentally alter
        the problem formulation, which makes DSM an excellent foundation for specifying the FPF itself. Traditional DSM only captures information 
        about data dependency between analyses. Objective and constraint information is missing from the description of the problem.

        An alternative matrix-based syntax, called a Functional Dependency Table (FDT), was proposed by Wagner and Papalambros. 
        FDT represents the relationship between functions, including objectives and constraints, and specific variables that affect 
        them\cite{Wagner1993}. Similar to DSM, FDT also describes an adjacency matrix of a graph. Unlike the DSM graph, 
        however, the graph is undirected and nodes can represent analysis tools, objectives, 
        or constraints. Edges between nodes represent a dependence on the same 
        variable. Michelena and Papalambros made use of the FDT to solve a graph partitioning problem that yielded 
        more efficient optimization problem decompositions\cite{Michelena1997}. While FDT succeeds at capturing the 
        information about objectives and constraints, it neglects the data dependency captured by DSM. For instance, 
        although we know from the FPF in Eq. \ref{eqn:simple_fpf} that the objective, $f$, is dependent on the 
        output of analysis $A$, you could not determine that from the FDT in Fig. \ref{fig:FDT_simple} alone. This make FDT 
        applicable for partitioning decisions, but not for ordering problems.

@@

        Lamb and Martins included the variables, objectives, and constraint functions as nodes in an Extended 
        DSM (XDSM) in order to capture a more complete description of solution strategies for MDAO problems\cite{Lambe2012}. By adding in those 
        elements, they partially combined a traditional DSM with an FDT. 
        This allows XDSM to represent data dependency between multiple analysis tools as well as between analysis tools and
        objective/constraint functions. With the additional information included in an XDSM, Lu and Martins applied both ordering and partitioning 
        algorithms on an MDAO test problem named the Scalable Problem \cite{Lu2012}.

@@

        By taking XDSM, removing the requirement for solver/optimizer nodes, and adding nodes for each variable, we are able to fully describe 
        the problem formulation for a general MDAO problem. In the following sections we describe the necessary syntax for specifying a 
        problem formulation within a DSM and show how to use the resulting graph to find the FPF from any given graph. For completeness we 
        also present algorithms for converting between the general DSM and the more specific formulations discussed above.

>>>>

    As the size and complexity of engineering systems grow, the time and expense for setting up 
    analysis models grow with them. Multidisciplinary Design Analysis and Optimization (MDAO)
    frameworks such as OpenMDAO\cite{Gray2012} and ModelCenter have enabled a new level of analysis tool integration 
    and paved the way for models with more analyses and increasing numbers of interdisciplinary couplings. That 
    new capability has created a new challenge since configuring a model with the larger nubmer of analyses, associated inputs and outputs
    can be very difficult. Even for models with 10's of analyses, there could be hundreds of variables that need to be managed. 
    It is not hard to imagine that the task of combining all the analyses into a consistent system model capable of solving a relevant 
    engineering design problem could easily become much more costly than createing the discipline analyses themselves. Thus, for a large system
    as the couplings between the disciplines begin to dominate the design space, the couplings between the analyses begin to dominate 
    the job of setting up the model.

    When building a model, coupling represents the reciprical flow of information between two analyses. 
    We propose the use of graph-based syntax to describe the flow of information from inputs, 
    through analyses, to outputs. If you allow for any free inputs to become design
    variables and some set of the outputs to become design objectives and constraints, then the graph 
    represents the problem formulation for a given design effort. 
    The graph-based syntax presented has application in all phases of the design process. 
    It can be used early on, to help construct a valid problem formulation from the set of available analysis tools. 
    It also provides a consistent structure to describe problem formulation that provides a foundation to apply
    algorithmic methods for selecting and then implementing effective solution strategies inside an MDAO framework.

    We've already identifed that a problem formulation graph should include design variables, analysis tools and their 
    associated inputs and outputs, and objectives and constraints. Notably absent from that list are any kind of 
    solvers, optimizers, or other iterative solution finding tools. At its most fundamental state, a problem formulation 
    includes only information about what is being sought after in a design or what the
    goals of a given problem are. It need not contain any information about solution paths or strategies to reach those goals. 
    While solution information can be represented with the graph syntax presented here, a unique aspect of this syntax is that 
    such information always remains seperable from the underlying problem formulation. In other words, a given a graph representing 
    a specific solution strategy for a specific problem formulation, it is possible to remove the solution information from the 
    graph and reclaim the more basic problem formulation.

\section{Specific vs Fundamental Problem Formulation }

    The Fundamental Problem Formulation (FPF) for any given problem will be constant regardless 
    of which MDAO framework, optimization algorithm, iterative solver, or solution strategy
    is used to solve the problem. For example, consider the following notional problem:

@@

    $A$ and $B$ represent analysis tools, and $f$ and $g$ are the objective and constraint functions respectively. 
    Equation \ref{eqn:simple} makes an inherent assumption about the solution strategy for the problem. 
    Analysis $A$ outputs $z$, which is an input to $B$. Hence, $A$ should be run before $B$ with 
    with $y$ being iterated on to convergence with $y^t$. However, a slightly different formulation is 
    equally valid and still represents the exact same problem:

@@

    Equation \ref{eqn:simple2} differs only slightly from Eq. \ref{eqn:simple}. $A$ is now dependent on the output of $B$, 
    and $z$ will be iterated on to convergence with $z^t$. Now the problem can be solved by running $B$ first and then $A$.
    Since the formulations in Eqs. \ref{eqn:simple} and \ref{eqn:simple2} both describe the same problem and neither can be the
    FPF. They are both specific versions of the same more fundamental description of 
    the problem that is common between them. We present the FPF as follows:

@@

\section{Existing Graph--Based Problem Formulation Syntax}

    As shown above, the mathematical language for specifying problem formulations is very general and can be used both for 
    fundamental and specific problem formulations. Tedford and Martins used the above mathematical syntax to specify the 
    FPF for a set of test problems and also to describe specific formulations for solving them with a 
    number of optimization architectures\cite{Tedford2009}. Their work demonstrates clearly how multiple specific 
    problem formulations can all relate back to a common FPF. The challenge with using this 
    traditional mathematical syntax is that it is not easily manipulated or analyzed. 
    A number of graph-based methods have been used successfully to translate the 
    mathematical syntax into a more useful computational form.

    Steward's Design Structure Matrix (DSM) is a square adjacency matrix which captures the relationship between analysis tools where off 
    diagonal elements of the matrix indicate coupling\cite{Steward1981}. Since a DSM describes a square adjacency matrix, 
    it can be represented in an equivalent directed graph where nodes represent analysis tools and 
    edges represent information dependence between those tools. The ordering of elements in a DSM can be used to indicate 
    execution order.  For more complex problems, choosing the proper order to run analysis tools is a non-trival task. 
    Rogers et. al developed DeMAID to manipulate a DSM to find an ordering for analysis tools that 
    reduces the cost of solving highly coupled systems\cite{Rogers1996}. This re-ordering is done through 
    row operations on the DSM matrix and yields multiple specific problem 
    formulations which all solve the same FPF. In other words, manipulation of the DSM does not fundamentally alter
    the problem formulation, which makes DSM an excellent foundation for specifying the FPF itself.

    Despite it's attractive properties, a DSM by itself is insufficient to describe complete problem formulations. 
    Traditional DSM only captures information about data dependency between analyses. 
    Objective and constraint information is missing from the description of the problem. 
    An alternative matrix-based syntax, called a Functional Dependency Table (FDT), was proposed by Wagner and Papalambros. 
    FDT represents the relationship between functions, including objectives and constraints, and specific variables that affect 
    them\cite{Wagner1993}. Similar to DSM, FDT also describes an adjacency matrix of a graph. Unlike the DSM graph, 
    however, the graph is undirected and nodes can represent analysis tools, objectives, 
    or constraints. Edges between nodes represent a dependence on the same 
    variable. Michelena and Papalambros made use of the FDT to solve a graph partitioning problem that yielded 
    more efficient optimization problem decompositions\cite{Michelena1997}. While FDT succeeds at capturing the 
    information about objectives and constraints, it can not capture the coupled data dependency that DSM captures. For instance, 
    we know from the FPF in Eq. \ref{eqn:simple_fpf} that the objective, $f$, is dependent on the 
    output, $m$, of analysis $A$. You could not determine that from the FDT in Fig. \ref{fig:FDT_simple} alone. This missing 
    information means that, while FDT is very useful for partitioning problems, it is also not sufficient to contain a 
    complete problem formulation.

@@

    Lamb and Martins included objectives and constraint functions as nodes in an Extended 
    DSM (XDSM) in order to capture a more complete description of solution strategies for MDAO problems\cite{Lambe2012}. 
    XDSM retains the square adjacency matrix form from DSM, but by adding in the 
    new elements they partially combined a traditional DSM with an FDT. 
    This allows XDSM to represent data dependency between multiple analysis tools as well as between analysis tools and
    objective/constraint functions. With the additional information included in an 
    XDSM, Lu and Martins applied both ordering and partitioning 
    algorithms on an MDAO test problem named the Scalable Problem \cite{Lu2012}.

@@

    Of the three methods, XDSM comes the closest to fully describing a problem formulation. 
    Thought it does not contain specific variable information, this could be easily added back in
    simply by putting the variables back into the graph and drawing the apprpriat connections
    between them and the relevant analysis nodes. However, it's dependence on 
    solution strategy specifics to represent couplings can not be removed without 
    stripping it of the useful information contained in an FDT. In the following sections
    we describe a new graph syntax that retains the features of XDSM, without the dependence on 
    solution strategy (although it does allow for that information to be present). Thus 
    the new syntax allows for the compelete specification of an FPF.
